Expression QTL hotspot dissection with mediation analysis and pleiotropy testing

### 2019-08-09 analysis/paper/Rmd/hotspot-one-chromosome.Rmd
R coding revisions

```{r pm-{{hot_chr}}}
pm <- pleio_tib %>%
  tibble_to_matrix(symmetric = TRUE)
```

@@

```{r, scols-{{hot_chr}}}
scols1 <- tibble(local = 1L + (rownames(pm) %in% local_annot2$symbol), 
                 symbol = rownames(pm), 
                 mediated_by_keller_mediator = rownames(pm) %in% (med_tib %>% filter(local_symbol == "{{keller_mediator}}", lod_diff > 1.5))$nonlocal_symbol
                 )

@@

```{r, ordering-{{hot_chr}}}
# order the nonlocal traits first
## indicator for being a nonlocal trait
indic_nonlocal <- rownames(pm) %in% nonlocal_annot$symbol
# define ord_nonlocal
ord_nonlocal <- pm[indic_nonlocal, ] %>%
  dist() %>%
  hclust() %>%
  as.dendrogram() %>%
  dendextend::seriate_dendrogram(dist(pm[indic_nonlocal, ])) %>%
  order.dendrogram()
rn_nonlocal <- rownames(pm[indic_nonlocal, ][ord_nonlocal, ])
# define ord_local
ord_local<- pm[!indic_nonlocal, ] %>%
  dist() %>%
  hclust() %>%
  as.dendrogram() %>%
  dendextend::seriate_dendrogram(dist(pm[!indic_nonlocal, ])) %>%
  order.dendrogram()
rn_local <- rownames(pm[!indic_nonlocal, ][ord_local, ])
rn <- c(rn_nonlocal, rn_local)
```

```{r, scols-peak-position-{{hot_chr}}}
annot_all <- local_annot2 %>%
  bind_rows(nonlocal_annot)

binning <- annot_all %>%
  dplyr::select(symbol, pos, lod) %>%
  mutate(bin = unlist(pos) %>%
           map(.f = function(x) {x %>% 
               bin_peak_position(breaks = hot_lower + (1:9) * 0.4) %>%
               binarize_bin_number(total_number_of_bins = 10L)
             })
  )
```

```{r, scan1-{{hot_chr}}}
fn <- here::here("analysis", "data", "derived_data", "Chr{{hot_chr}}_scan1_out.rds")
if (file.exists(fn)) {
  s1out <- readRDS(fn)
} else {
  s1out <- qtl2::scan1(genoprobs = probs, 
                       pheno = cbind(hot_local, hot_nonlocal), 
                       kinship = K[["{{hot_chr}}"]], 
                       addcovar = covar, 
                       reml = TRUE
                       )
  saveRDS(s1out, fn)
}
```

```{r allele-effects-{{hot_chr}}}
fn <- here::here("analysis", "data", "derived_data", "Chr{{hot_chr}}_scan1coef_out.rds")
if (file.exists(fn)) {
  s1coef_out <- readRDS(fn)
} else {
s1coef_out <- purrr::map(.x = tibble(cbind(hot_local, hot_nonlocal)), 
           .f = function(pheno) {
             qtl2::scan1coef(genoprobs = probs[["{{hot_chr}}"]], 
                             pheno = pheno, 
                             kinship = K[["{{hot_chr}}"]], 
                             addcovar = covar, 
                             reml = TRUE, 
                             )
           }
                )
saveRDS(s1coef_out, fn)
}
```

```{r, get-allele-effects-{{hot_chr}}}
# add index values for peak position
local_annot2 %>%
  mutate(peak_index = match(table = names(map[["{{hot_chr}}"]]), x = marker.id)) %>%
  mutate(peak_effects = qtl2pleio::get_effects(peak_index, allele_effects_matrix = s1coef_out, map = map[["{{hot_chr}}"]]))

nonlocal_annot %>%
  mutate(peak_index = match(table = names(map[["{{hot_chr}}"]]), x = marker.id))

```{r define-scols-pre-{{hot_chr}}}
scols_pre <- scols_extra %>%
  dplyr::select(n_med_per_local$local_symbol)
# get ordering for columns of scols_pre
col_order <- scols_pre %>%
  t() %>%
  dist() %>%
  hclust() %>%
  as.dendrogram() %>%
  dendextend::seriate_dendrogram(dist(t(scols_pre))) %>%
  order.dendrogram()
scols <- scols_pre[ , col_order] %>%
  mutate(local = scols1)
```

```{r, heat-{{hot_chr}}, fig.cap="Heatmap for pleiotropy test statistics for Chr {{hot_chr}}."}
ord <- match(rn, rownames(pm))

(p <- pm[ord, ord] %>% # order both cols and rows by 'ord'
  apply(FUN = trunc2, MARGIN = 2, threshold = 5) %>%
  heatmaply::heatmapr(Rowv = FALSE,
            Colv = FALSE,
            plot_method = "plotly"
            ) %>%
    heatmaply::heatmaply_heatmapr2(
            fontsize_col = 2,
            fontsize_row = 2,
            symm = TRUE,
            row_side_colors = list(scols[ord, ], tibble(scols1$local[ord])),
            row_side_palette = list(heatmaply::cool_warm, heatmaply::Blues), 
            col_side_colors = tibble(local = scols1$local[ord]),
            limits = c(0, 5)
            ) %>%
  hide_guides()
)
```

>>>>

```{r pm-{{hot_chr}}}
pm <- pleio_tib %>%
  arrange(gene1_symbol, gene2_symbol) %>%
  tibble_to_matrix(symmetric = TRUE)
```

@@

```{r, scols-{{hot_chr}}}
scols1 <- tibble(local = 1L + (rownames(pm) %in% local_annot2$symbol), 
                 symbol = rownames(pm), 
                 mediated_by_keller_mediator = rownames(pm) %in% (med_tib %>% filter(local_symbol == "{{keller_mediator}}", lod_diff > 1.5))$nonlocal_symbol
                 )
```

```{r, ordering-{{hot_chr}}}
# order the nonlocal traits first
## indicator for being a nonlocal trait
indic_nonlocal <- rownames(pm) %in% nonlocal_annot$symbol
# define ord_nonlocal
ord_nonlocal <- pm[indic_nonlocal, ] %>%
  dist() %>%
  hclust() %>%
  as.dendrogram() %>%
  dendextend::seriate_dendrogram(dist(pm[indic_nonlocal, ])) %>%
  order.dendrogram()
rn_nonlocal <- rownames(pm[indic_nonlocal, ][ord_nonlocal, ])
# define ord_local
ord_local<- pm[!indic_nonlocal, ] %>%
  dist() %>%
  hclust() %>%
  as.dendrogram() %>%
  dendextend::seriate_dendrogram(dist(pm[!indic_nonlocal, ])) %>%
  order.dendrogram()
rn_local <- rownames(pm[!indic_nonlocal, ][ord_local, ])
rn <- c(rn_nonlocal, rn_local) # order for dendrogram
```

```{r, scols-peak-position-{{hot_chr}}}

binning <- local_nonlocal_annot %>%
  dplyr::select(symbol, pos, lod) %>%
  mutate(bin = unlist(pos) %>%
           map(.f = function(x) {x %>% 
               bin_peak_position(breaks = hot_lower + (1:9) * 0.4) %>%
               binarize_bin_number(total_number_of_bins = 10L)
             })
  ) %>%
  listcol_to_cols(list_column = quo(bin), another_column = quo(symbol))
```

```{r, scan1-{{hot_chr}}}
fn <- here::here("analysis", "data", "derived_data", "Chr{{hot_chr}}_scan1_out.rds")
if (file.exists(fn)) {
  s1out <- readRDS(fn)
} else {
  s1out <- qtl2::scan1(genoprobs = probs, 
                       pheno = cbind(hot_local, hot_nonlocal), 
                       kinship = K[["{{hot_chr}}"]], 
                       addcovar = covar, 
                       reml = TRUE, 
                       cores = 7
                       )
  saveRDS(s1out, fn)
}
```

```{r allele-effects-{{hot_chr}}}
fn <- here::here("analysis", "data", "derived_data", "Chr{{hot_chr}}_scan1coef_out.rds")
if (file.exists(fn)) {
  s1coef_out <- readRDS(fn)
} else {
s1coef_out <- purrr::map(.x = as_tibble(cbind(hot_local, hot_nonlocal)), 
           .f = function(pheno) {
             ph <- as.matrix(pheno)
             rownames(ph) <- rownames(expr)
             qtl2::scan1coef(genoprobs = list(probs[["{{hot_chr}}"]]), 
                             pheno = ph, 
                             kinship = K[["{{hot_chr}}"]], 
                             addcovar = covar, 
                             reml = TRUE, 
                             cores = 7
                             )
           }
                )
saveRDS(s1coef_out, fn)
}
```

```{r, get-allele-effects-{{hot_chr}}}
# add index values for peak position
local_nonlocal_annot <- dplyr::bind_rows(local_annot2, nonlocal_annot) %>%
  mutate(peak_index = match(table = names(map[["{{hot_chr}}"]]), x = marker.id)) %>%
  mutate(peak_effects = purrr::map2(.f = qtl2pleio::get_effects, .x = peak_index, .y = s1coef_out, map = map[["{{hot_chr}}"]])) %>%
  listcol_to_cols(list_column = quo(peak_effects), another_column = quo(symbol))
```

```{r define-scols-pre-{{hot_chr}}}
scols_pre <- scols_extra %>%
  dplyr::select(n_med_per_local$local_symbol)
# get ordering for columns of scols_pre
col_order <- scols_pre %>%
  t() %>%
  dist() %>%
  hclust() %>%
  as.dendrogram() %>%
  dendextend::seriate_dendrogram(dist(t(scols_pre))) %>%
  order.dendrogram()
scols <- scols_pre[ , col_order] %>%
  mutate(local = scols1$local)
```

```{r, prepare-annotations-{{hot_chr}}}
# need 5 annotation object: 1. local v nonlocal, 2. mediation LOD difference for all local traits, 3. mediation LOD difference proportion for all local traits, 4. 8 allele effects per trait, 5. peak position per trait, encoded as binned position
# 2,3,4 are numeric; 1 & 5 are binary
## local_nonlocal_annot already has the allele effect and local v nonlocal; binning has the peak position. Need to join these two on symbol
### med_tib has lod diff & lod diff proportion for all local - nonlocal pairs
fake_tibble_mediation <- tidyr::crossing(symbol = local_annot2$symbol, 
                                         symbol2 = local_annot2$symbol) %>%
  mutate(dat = NA) %>%
  spread(key = symbol, value = dat) %>%
  rename(symbol = symbol2)

lod_diffs_tibble <- med_tib %>%
  select( - c(lod_without_med, 
              lod_with_med, 
              lod_diff_proportion, 
              local_gene_id, 
              nonlocal_gene_id)
          ) %>%
  tidyr::spread(key = local_symbol, 
                value = lod_diff) %>% 
  dplyr::rename(symbol = nonlocal_symbol) %>%
  dplyr::bind_rows(fake_tibble_mediation)
lod_diff_proportions_tibble <- med_tib %>%
  select( - c(lod_without_med, 
              lod_with_med, 
              lod_diff, 
              local_gene_id, 
              nonlocal_gene_id)
          ) %>%
  tidyr::spread(key = local_symbol, 
                value = lod_diff_proportion) %>% # 147 rows by 32 columns
  rename(symbol = nonlocal_symbol) %>%
  bind_rows(fake_tibble_mediation)
allele_effects_tibble <- local_nonlocal_annot %>%
  dplyr::select(symbol, 18:25)
peak_positions <- binning %>%
  select(symbol, pos)
local_tibble <- local_nonlocal_annot %>%
  select(symbol, local_tx) %>% 
  ungroup()

@@

```{r, heat-{{hot_chr}}, fig.cap="Heatmap for pleiotropy test statistics for Chr {{hot_chr}}."}
ord <- match(x = rn, table = rownames(pm))

pm[ord, ord] %>% # order both cols and rows by 'ord'
  apply(FUN = trunc2, MARGIN = 2, threshold = 5) %>%
  heatmaply::heatmapr(Rowv = FALSE,
            Colv = FALSE,
            plot_method = "plotly"
            ) %>%
    heatmaply::heatmaply_heatmapr2(
            fontsize_col = 2,
            fontsize_row = 2,
            symm = TRUE,
            row_side_colors = list(scols[ord, ], tibble(scols1$local[ord])),
            row_side_palette = list(heatmaply::cool_warm, heatmaply::Blues), 
            col_side_colors = tibble(local = scols1$local[ord]),
            limits = c(0, 5)
            ) %>%
  hide_guides()
```

### 2019-08-09 analysis/paper/Rmd/hotspots-paper.Rmd
removed unused assignment
updated paper.Rmd
R coding revisions

```{r setup}
library(knitr)
library(qtl2hotspots)
pleio_threshold <- 4
ult <- 15 # univariate lod threshold
```

@@

```{r readRDS}
lod_peaks <- readRDS(here("analysis", "data", "derived_data", "lod_peaks.rds"))
annots <- readRDS(here("analysis", "data", "derived_data", "annots.rds"))
expr <- readRDS(here("analysis", "data", "derived_data", "expr.rds"))
```

```{r, load_purrr}
library(purrr)
```

@@

### Chr 2 hotspot results

@@

### Chr 5 hotspot results

@@

### Chr 7 results

@@

### Chr 11 results

@@

### Chr 13 results

>>>>

```{r setup}
library(knitr)
library(qtl2hotspots)
ult <- 15 # univariate lod threshold
```

@@

```{r readRDS}
lod_peaks <- readRDS(here("analysis", "data", "derived_data", "lod_peaks.rds"))
annots <- readRDS(here("analysis", "data", "derived_data", "annots.rds"))
expr <- readRDS(here("analysis", "data", "derived_data", "expr.rds"))
probs <- readRDS(here("analysis", "data", "derived_data", "genoprobs.rds"))
K <- readRDS(here("analysis", "data", "derived_data", "kinship.rds"))
covar <- readRDS(here("analysis", "data", "derived_data", "covar.rds"))
map <- readRDS(here("analysis", "data", "derived_data", "map.rds"))
```

@@

@@

@@

@@

@@
